%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter*{Danke sch\"on!}
\addcontentsline{toc}{chapter}{Danke sch\"on!}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

It was a cold and rainy night, in the first week of February. I was tired. No, I
was exhausted. After several hours flying from Brazil, I missed the last train
connection to Bielefeld because of a delay in my flight to Frankfurt and had to
wait a few hours, alone, in the train station in Cologne. I was carrying a lot
of luggage. A homeless old man looked puzzled at me. It was freezing, and I
quickly realized I would need to buy some \emph{real} winter clothes soon. When
I finally arrived in Bielefeld, it was already half past five in the morning.
The taxi driver couldn't speak English, but I showed him the hotel's name on a
crushed piece of paper. In the hotel room, I felt that what-am-I-doing-here
feeling once again in my life.

My fears of having made a terrible mistake vanished as soon as I knocked on
Jens' office. I was warmly welcome by him and Heike. Micha took me in his car
and helped me move my stuff from the hotel to the apartment where I would spend
the next few weeks. In the following days, I met a string of exceptionally
bright and nice colleagues: Conni, Gregor, Kim, Klaus, Mitch, Nando, Rileen,
Sebastian, Sven, Thomas, and Zsuzsa, just to name a few of the
AGGI group members at that time. In following years, some members left, others
joined, but all of them have made, somehow, a good impression on me: Anton,
Inke, Julia, Katharina, Nondas, Peter, Veli, and Wiebke are just a few
of the more recently members, with whom I have enjoyed lunch, coffee-breaks,
trips and night-outs innumerous times. At the university, I also met truly great
people from the GK Bioinformatik and the Graduate School of Bioinformatics:
Abhishek, Britta, Dirk Evers, Hari, Hiro, Janina, Julia Herold, Kostas, Lutz,
Naryttza, Ning Wei, Robert Heinen, Robert Giegerich, Roland, Sebastian Oehm,
Silke, Thomas Hoechsmann, Vijay, and many others.

I am also fortunate to have made a few good long-lasting friends in Bielefeld:
Zsuzsa, R\'eka, No\'emi and Nando; Anja, Niklas and Klaus; Monica, Marina and Nils;
Leo and Gisele; Poti and Helena; Jomuna and Jannik; Simone, Dani, Chico, Z\'e, and
many others from a lively (and fast-growing) Brazilian community who have helped
me feel a little at home on this side of the planet.

From the other side of the planet I received constant support and incentive from
family and friends that make me tremendously proud. Everything I have (and
everything I am) I owe to my parents, S\'ergio and Sandra, who have built the
most beautiful and loving family one can wish. My grandmother Lenise, my sisters
Juliana and Cl\'audia, my brothers-in-law Sparano and Julio, my nephews Felipe
and Gabriel, and my niece Victoria, all deserve heartfelt thanks.

But, if there is someone who deserves all the credit, she is my wife, Karla. She
embarked on this trip with me. She transcended a number of barriers. And she
succeeded. She continues to surprise me with every little accomplishment that
she makes. With remarkable energy, grace and infinite love, she made it
possible.

\vspace*{6ex}
S\'ergio A. de Carvalho Jr.\hfill Bielefeld, \finalmonth
